---
title: "Intro to R - Day 2"
author: "Daniel Fuller"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gtsummary)
library(vtable)
library(readxl)
library(reportRmd)
library(infer)
```

## Intro to R Day 2

### Importing data set in CSV file named `data1`

Make sure you check your file extension. If you have an excel file use `readxl` package or the `GUI` 

#### Tidyverse method

```{r}
data1 <- read_csv("data1.csv")

## Difference between Markdown and running code
#data1 <- read_csv("data_code/data1.csv") 
```

#### Base R method

```{r}
data1_base <- read.csv("data1.csv", header = TRUE)
```

#### Excel file

```{r}
data1_excel <- read_excel("data1.xlsx")
```

### to see first 10 observations

```{r}
head(data1,10)
```

### Importing data set in CSV file named ‘data2’

```{r}
data2 <- read_csv("data2.csv")
```

### Merging data1 and data2

First we want to create two artificial datasets to merge. We are just going to select a few variables from each then put the data back together.

```{r}
data1 <- dplyr::select(data1, id, sex, ethgrp, weight, age, cvd)
data2 <- dplyr::select(data2, id, stroke, smoking, Cancer, ldl1, ldl2, gender)
```

### Joining data 

In computer science the term is join. Some stat software uses the term merge. 

```{r}
data_merge <- dplyr::full_join(data1, data2) 

data_merge1 <- dplyr::full_join(data1, data2, by = join_by(id))

data_merge2 <- dplyr::full_join(data1, data2,by = join_by(id == id))
```

### importing test data

```{r}
test <- read_csv("test.csv")
head(test,10)
glimpse(test)

st(test) ## vtable package
rm_covsum(data = test, 
          covs = c('age', 'sex', 'stroke', 'Cancer', 'cvd'),
          show.tests=TRUE)         ## reportRmd package
```

### Create categorial variable ‘agecat’ using age

Here we are introducing three new things `%>%` (`pipe operator`), `mutate`, and `case_when`

* `%>%` or `|>` (`pipe operator`) = Signifies to run the analysis and move down to the next function. 
* `mutate` = The verb for create a new variable from an old variable
* `case_when`= An `if_else` type function 

```{r}
### Tidyverse Method
test |> summary(age)

### Base R Method
summary(test, age)
```

```{r}
test <- test %>% 
          mutate(age_cat = case_when(
            age < 45 ~ "<45",
            age >= 45 & age < 50 ~ "45-49",
            age >= 50 & age < 59 ~ "50-59",
            age >= 60 & age < 65 ~ "60-64",  
            TRUE ~ "65+"
          ))
```

```{r}
### Tidyverse Method
test %>% count(age_cat)

### Base R Method
count(test, age_cat)
```

### Writing a test to see if that worked

```{r}
### Tidyverse Method
#test %>% table(age, age_cat)

### Base R Method
table(test$age, test$age_cat)
```

### Check frequency distribution of a categorical variable

#### Check frequency distribution of gender

```{r}
table(test$gender)
```

#### Cross tabulation of gender and stroke

```{r}
table(test$gender, test$stroke)
```

#### Chi-square test between gender and stroke

Lots of the previous existing packages work with traditional coding rather than Tidyverse style. Here we are going to work on some of those older methods. If you want to use the newer versions use the [infer](https://infer.tidymodels.org/index.html) package. 

```{r}
### Tidyverse
test$stroke_factor <- as.factor(test$stroke)
chisq_test(test, gender ~ stroke_factor)

### Base R
chisq.test(test$gender, test$stroke)
```

# Fishers exact test

```{r}
fisher.test(test$gender, test$stroke)
```

#### First, we check if data follow Normal distribution
#### Normality test (if p>0.05: data are normally distributed)
####  Want to compare mean ages between male and female

####  Check histogtam
```{r}
hist_age_gender <- ggplot(test, aes(age)) +
  geom_histogram() + 
  facet_wrap(~ gender)

plot(hist_age_gender)
```

#### Shapiro test females

```{r}
test_data_female <- filter(test, gender == "f")
shapiro.test(test_data_female$age)
```

#### Shapiro test males
```{r}
test_data_male <- filter(test, gender == "m")
shapiro.test(test_data_male$age)
```

### t-test

Lots of things we can do with t-tests. Generic formalution for many tests is `test_function`(`variable1` ~ `variable1`, data = `my_data`)

```{r}
help(t.test)
t.test(age ~ gender, data = test, var.equal = FALSE)
```

### Check variance

```{r}
var.test(age ~ gender, data = test)
```

### Variances are equal based on the test
```{r}
t.test(age ~ gender, data = test, var.equal = TRUE)
```

### Non-parametric test if data are not normally distributed

```{r}
wilcox.test(age ~ gender, data = test)
```

### Paired t-test for two dependent samples

```{r}
t.test(test$ldl1, test$ldl2, paired = TRUE, var.equal = FALSE)
```

### Pearson correlation coefficient
```{r}
cor.test(test$ldl1, test$ldl2,  method = "pearson") 
```

### Spearman correlation coefficient
```{r}
cor.test(test$ldl1, test$ldl2,  method = "spearman") 
```

## Linear regression

Conduct linear regression model between dependent and independent variables. Age is continuous, gender is categorical, ldl is continuous variable
```{r}
linear_model <- lm(age ~ as.factor(gender) + ldl1, data = test)
summary(linear_model)

gtsummary::tbl_regression(linear_model)
```

## Logistic regression

```{r}
logistic_model <- glm(Cancer ~ as.factor(gender) + ldl1 + smoking, data = test, family = "binomial")
summary(logistic_model)
```

### Get ORs

#### Old School Way
```{r}
exp(cbind(OR = coef(logistic_model), confint(logistic_model)))

```

#### New School Way
```{r}
tbl_regression(logistic_model, exponentiate = TRUE) 
```

### One-way ANOVA

Pass arguments to aov() function for an ANOVA test

```{r}
one_anova <- aov(age ~ ethgrp, data = test)
summary(one_anova)
```

### Non-parametric ANOVA

```{r}
kruskal.test(age ~ ethgrp, data = test)
```