---
title: "Intro to R Day 3"
output:
      html_document:
        keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gtsummary)
library(sjPlot)
library(interactions)
library(cardx)
library(mice)
library(nlme)
library(randomNames)
```

## Data Wrangling 

```{r}
data <- read_csv("data1.csv")

cols <- c("id", "sex", "ethgrp", "cvd", "stroke", "smoking", "Cancer")
data <- data %>% mutate_at(cols, factor)
str(data)
```

### Joining datasets

#### Stacking 2 datasets on top of each other

If we want to stack two datasets we need to use the `bind` set of functions. We can either use `bind_rows` or `bind_cols` from the `tidyverse` set of packages to do this. Let's say we have two datasets with the same information but different participants. We want to stack them so we can do some bigger analysis. 

```{r}
data1 <- read_csv("data1.csv")
data2 <- read_csv("data2.csv")
```

Data1 and Data2 have the some number and information in the columns but they are not together. We can join them using `bind_rows`. 

```{r}
data_1_2 <- bind_rows(data1, data2)
```

We can also do more advanced things like select and bind all of the files in a folder with a specific filename or file extension. 

```{r}
paths <- list.files(pattern = "[.]csv$", full.names = TRUE)
paths

files <- map(paths, read_csv)
length(files)

data_all <- list_rbind(files)
#write_csv(data_all, "data_all.csv")
```

2. Join by ID




## Data Viz

### 1. ggplot2 General

ggplot2 is an open-source data visualization package for the statistical programming language R. Created by Hadley Wickham in 2005, ggplot2 is an implementation of Leland Wilkinson's Grammar of Graphics—a general scheme for data visualization which breaks up graphs into semantic components such as scales and layers.[Wikipedia](https://en.wikipedia.org/wiki/Ggplot2) The basic idea is to build plots layer by layer. With ggplot2 you can control absolutely everything about a plot. 

Data viz is a big place where R and Stata diverge a lot. With R giving much more flexibility (and sometimes annoyance) in terms of visualizing data. We are going to cover basics but recommend the BBC Code Book [https://bbc.github.io/rcookbook/](https://bbc.github.io/rcookbook/) and Keiran Healey's book _Data Visualization_ [https://kieranhealy.org/publications/dataviz/](https://kieranhealy.org/publications/dataviz/) which is part of the the `socviz` package [https://kjhealy.github.io/socviz/](https://kjhealy.github.io/socviz/. )

### 2. Histograms and single variable box plots

ggplot2 is built on things called *geoms* which represent different types of plots. There are *geoms* for histograms, scatterplots, beeswarm, and many other plots. Here we are focusing two basic plots and the defaults of everything else. 

#### Single bar graphs

People get stuck in R making bar graphs because they are used to Excel and only have the mean value of a given column then making the bar graph from that mean column. It's always questionnable whether you should be making a bar graph, but if you really need to here is how you can do it in R. 

```{r}
# Recoding income here so we can see the levels more easily

data_all <- data_all %>%
	mutate(stroke_recode = case_when(
		stroke == "1" ~ "Yes",
    stroke == "0" ~ "No"
	))

data_all$stroke_recode <- as_factor(data_all$stroke_recode)

table(data_all$stroke, data_all$stroke_recode)
```

```{r}
bar_stroke <- ggplot(data_all, aes(stroke_recode)) + 
                geom_bar() + 
                labs(x = "Has had a stroke") + 
                theme(axis.text.x = element_text(angle = 90))       
plot(bar_stroke)
```

#### Single variable boxplots

```{r}
boxplot <- ggplot(data_all, aes(stroke_recode)) + 
              geom_boxplot() + 
              coord_flip()  # Here we add coord_flip function to make the boxplot more as we would expect
plot(boxplot)
```

#### Two variable boxplots

Very useful for showing mean differences and presenting stuff with ANOVA type analysis. 

```{r}
boxplot_stroke <- ggplot(data_all, aes(x = age, y = stroke_recode, colour = stroke_recode)) + 
              geom_boxplot() + 
              coord_flip() + # Here we add coord_flip function to make the boxplot more as we would expect
              theme_classic()
plot(boxplot_stroke)
```

### 3. Scatter plots

Scatterplots plot the relationship between two variables. There are lots of things we can do and we will build a plot sequentially. We are going to plot the relationship between age and physical activity (two continuous variables). 

```{r}
scatter_plot <- ggplot(data_all, aes(x = age, y = ldl1)) + 
                  geom_point(position=position_jitter())
plot(scatter_plot)
```

Common things you will see with a scatter plot including the following

#### Adding a regression line

Here we add another *geom* on top of the *geom_point* to start building our plot. We will use *geom_smooth* to add a line. The default in R is a lowess smoother. You can also add a linear regression line. I'm also changing the colour so we can see the difference between the two lines. It's best to use hexcodes for colours but we can also just use words. 

```{r}
scatter_plot_line <- ggplot(data_all, aes(x = age, y = ldl1)) + 
                  geom_point(position = position_jitter()) + 
                  geom_smooth(colour = "red") + 
                  geom_smooth(method = "lm", colour = "#088da5")
plot(scatter_plot_line)
```

#### Changing the variable names

```{r}
scatter_plot_variables <- ggplot(data_all, aes(x = age, y = ldl1)) + 
                  geom_point(position = position_jitter()) + 
                  geom_smooth(colour = "red") + 
                  geom_smooth(method = "lm", colour = "#088da5") +
                  labs(x = "Age", y = "LDL Cholesterol")
plot(scatter_plot_variables)
```

#### Changing the shading of the points

We can use *alpha* to change the shading of the points. This lets use quickly avoid overplotting with lots of overlapping points. We need to play with this a bit. A good place to start is 0.5 and go from there. I ended up at 0.2 which I think shows the mass of data points and avoids too much emphasis on the outlier points. 

```{r}
scatter_plot_alpha <- ggplot(data_all, aes(x = age, y = ldl1)) + 
                  geom_point(alpha = 0.2, position = position_jitter()) + 
                  geom_smooth(colour = "red") + 
                  geom_smooth(method = "lm", colour = "#088da5") +
                  labs(x = "Age", y = "LDL Cholesterol")
plot(scatter_plot_alpha)
```

### 4. Grouping with ggplot2

One of the best things about ggplot2 is the ability to easily *group_by* like we would do with data wrangling. We do this by adding groupings (colouring by a variable) or facets (creating separate plots). Here we want to group by gender to see if there are visual differences between genders in the age-PA association. 

Colouring by gender
```{r}
scatter_plot_gender <- ggplot(data_all, aes(x = age, y = ldl1, colour = stroke_recode)) + 
                  geom_point(alpha = 0.2) + 
                  labs(x = "Age", y = "LDL Cholesterol", fill = "Gender") 
plot(scatter_plot_gender)
```

Faceting by gender
```{r}
scatter_plot_gender <- ggplot(data_all, aes(x = age, y = ldl1)) + 
                  geom_point(alpha = 0.2) + 
                  labs(x = "Age", y = "LDL Cholesterol") +
                  facet_wrap(~ stroke_recode)
plot(scatter_plot_gender)
```

### 5. Colours and themes

Finally, there are many default colours and themes we can use to make the plots look good very quickly. A few things I use regularly. 

#### Themes

There are two themes I use regularly; *classic* and *bw*. Classic provides a clean looking plot with no/limited background. *bw* provides a black and white figure, which is great for publications that need no colour. 

**Black and White Theme**

```{r}
scatter_plot_bw <- ggplot(data_all, aes(x = age, y = ldl1, colour = stroke_recode)) + 
                  geom_point(alpha = 0.2) + 
                  labs(x = "Age", y = "LDL Cholesterol", colour = "Gender") +
                  theme_bw()
plot(scatter_plot_bw)
```

**Classic Theme**

```{r}
scatter_plot_classic <- ggplot(data_all, aes(x = age, y = ldl1, colour = stroke_recode)) + 
                  geom_point(alpha = 0.2) + 
                  labs(x = "Age", y = "LDL Cholesterol", colour = "Gender") +
                  theme_classic()
plot(scatter_plot_classic)
```

#### Colours

There are lots of default colours in R. The basic functions are *scale_colour* and *scale_fill*. We didn't talk about this much but colour and fill are different based what you are filling. If we go back to the gender colouring we can use different functions to change the colours. The colour brewer colours [https://colorbrewer2.org/](https://colorbrewer2.org/) using *scale_colour_brewer* or *scale_fill_brewer* for sensible default colours. 

```{r}
scatter_plot_gender_brewer <- ggplot(data_all, aes(x = age, y = ldl1, colour = stroke_recode)) + 
                  geom_point(alpha = 0.2) + 
                  scale_colour_manual(values = c("#E69F00", "#56B4E9")) +
                  labs(x = "Age", y = "LDL Cholesterol", colour = "Gender") 
plot(scatter_plot_gender_brewer)
```

### 6. Heat Maps 

Someone asked about heat maps so here is an example. A heatmap is a row and column look at the data so we need to do some reshaping and selecting the data in order for this to work. We are only going to use age, ldl1, and ldl2 for this to keep things simple. 

```{r}
data_heat <- data_all %>% select(id, ldl1, ldl2)


data_heat <- data_heat %>%
                dplyr::mutate(id = row_number()) %>%
                gather(-id, key = "key", value = "val") %>%
                mutate(isna = is.na(val)) 

ggplot(data = data_heat, aes(id, key, fill = val)) +
    geom_tile(alpha=1) +
    labs(x = "Person", y = "LD 1 and 2") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 
```

## Regression stuff 

### Creating a table of bivariable associations

We always want to look at all of the bivariate associations for each independent variable. We can do this quickly with the final fit package. For now ignore the multivariable model results. We just want to look at the bivariable. 

```{r}
univ_table <- data_all %>%
  select(ldl1, sex, ethgrp, age, smoking, Cancer, stroke_recode) %>%
  tbl_uvregression(
    method = glm,
    y = ldl1,
    method.args = list(family = gaussian)) 

univ_table %>% as_kable()
```

### Creating a final model

Here we can create a final model and plot the model results along with our 

```{r}
model_final <- glm(ldl1 ~  sex +
                           ethgrp + 
                           age + 
                           smoking + 
                           Cancer + 
                           stroke_recode, 
                    data = data_all, family = "gaussian")

summary(model_final)

multi_table <- tbl_regression(model_final) 

multi_table %>% as_kable()
```

### Plot a final model

We might also want to plot the results of the final model as coefficients and confidence intervals instead of showing the model in table. Can be a nice way to present the results. We are going to use the `plot_model` from the `sjPlot` package. This package leverages `ggplot2` but probably goes not offer quite as much flexibility. 

```{r}
plot_model(model_final, type="est")
```

## Interaction terms

We might want to check for interaction terms. There might be an interaction between smoking and cancer so let's test that our dataset, visualize the interaction, and compare to both our bivariable models and final model. 

```{r}
model_interaction <- glm(ldl1 ~  sex +
                           ethgrp + 
                           age + 
                           smoking + 
                           Cancer + 
                           smoking*Cancer + 
                           stroke_recode, 
                    data = data_all, family = "gaussian")

summary(model_interaction)

inter_table <- tbl_regression(model_interaction) 

inter_table %>% as_kable()
```

### Visualize interaction term

```{r}
plot_interaction <- interact_plot(model_interaction, 
                                    pred = smoking, 
                                    modx = Cancer, 
                                    plot.points = TRUE, 
                                    partial.residuals = TRUE, 
                                    interval = TRUE,
                                    int.width = 0.95)
plot(plot_interaction)
```

## Putting it all together

We might want to make a big table with our bivariable analysis, final model, and interaction model for presentation or quickly show results. We can do that by combining tables

```{r}
tbl_univ_multi_inter <- tbl_merge(list(univ_table, multi_table, inter_table))

tbl_univ_multi_inter %>% as_kable()
```

For linear regression models there are a number of assumptions we want to check. These are typically easy visualized using plots. We can produce the plots for the relevant visualizations using `plot` from base R or by using 

```{r}
plot(model_final)

#### Include the TidyR for this. 
```

## Multilevel Models



```{r, message = FALSE, warning = FALSE}
ldl1_age <- ggplot(data_all, aes(x = age, y = ldl1)) + 
                    geom_point() + 
                    geom_smooth(method='lm', formula= y~x)
plot(ldl1_age)
```

## Variation between height and weight by health region

```{r, message = FALSE, warning = FALSE}
age_ldl1_parent <- ggplot(data_all, aes(x = age, y = ldl1, colour = parent, fill = parent)) + 
                    geom_point() + 
                    geom_smooth(method='lm', formula= y~x) + 
                    theme(legend.position="none")
plot(age_ldl1_parent)
```

## Modelling intercepts

Linear regression. Assumes only one intercept for the entire sample. 

```{r}
lm_1 <- lm(ldl1 ~ 1, data = data_all)
```

Multil-level model. Assumes one intercept for each group. In our case parents, it could be people, neighbourhoods, etc. 

```{r}
lme_1 <- lme(ldl1 ~ 1, random = ~ 1 | parent, data = data_all)
summary(lme_1)
```

1. Fixed effects. Our level 1 variables go here. In this case the ~ 1 means model only the intercept.  
2. Random effects. The intercept is computed at the level of health region (nesting factor), and that health region is treated as a random variable.

```{r}
lme_1 <- lme(ldl1 ~ 1, random = ~ 1 | parent, data = data_all)


intercepts <- lme_1[["coefficients"]][["random"]][["health_region"]] + lme_1[["coefficients"]][["fixed"]][["(Intercept)"]]
sort(intercepts) 
```

## Plot the distribution of intercepts

```{r}
ldl1_parent <- ggplot() + 
              geom_histogram(aes(intercepts))
plot(ldl1_parent)
```

## Adding level 1 covariates

```{r}
lme_2 <- lme(ldl1 ~ age + stroke_recode + smoking, random = ~ 1 | parent, data = data_all)
summary(lme_2)
```

## Plotting fitted versus actual values

```{r}
data_all$fitted_lm <- predict(lme_2)

cchs_small <- subset(cchs, geodpmf < 12000)

fitted_plot <- ggplot(data_all, aes(x = age, y = ldl1, colour = parent)) + 
                        geom_point() + 
                        geom_line(aes(y = fitted_lm, colour = parent)) +
                        theme(legend.position="none") 
plot(fitted_plot)
```

### Code for creating synthetic datasets

Not run

```{}
where <- make.where(data, "all")

method <- make.method(data, where = where)
method[method == "pmm"] <- "norm"

syn_param <- mice(data, 
                  m = 10, 
                  maxit = 1,
                  method = "cart",
                  where = where,
                  printFlag = FALSE)

data2 <- complete(syn_param, 1)
data3 <- complete(syn_param, 2)
data4 <- complete(syn_param, 3)

write.csv(data2, "data2.csv")
write.csv(data3, "data3.csv")
write.csv(data4, "data4.csv")
```

### Generating random names

```{}
names_data <- NULL

names_data$names <- randomNames(800, 
                      which.names="first", 
                      sample.with.replacement=TRUE,
                      return.complete.data=FALSE) 
names_data$id <- seq(from = 1, to = 800)

names_data <- as.data.frame(names_data)

write.csv(names_data, "names.data.csv")
```


